Relationships in the extremes and their influence on competition and coexistence

### 2022-04-27 paper/Paper.Rnw
spelling

%G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decomponse
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

@@

%ATA specialists
Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist in some contexts. If a strong competitor is present across a
collection of habitat patches and the GWR of a weaker competitor is only positive 
in a patch in the presence of 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present.
The weaker competitor can therefore perhaps be regarded as an ``ATA specialist'' in the landscape. EC-associations
can often boil down to associations between two environmental variables, as was assumed to be the case
for the lottery model. Mechanisms of coexitence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs are absent, and is hence a specialist on scenarios with symmetric
tail associations of environmental variables (an ``STA specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

@@

%E and vital rates
Whereas the ``$E$'' of ``$EC$-covariance'' is often conceptualized as ``environment'', in fact it is a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to actual environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is actually better conceptualized as a species response
than as the environment, directly.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, clearly a response
to the environment rather than a direct environmental measurement. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
Thus the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population responses which actually matter for coexistence.
Vital rates are often measured directly in demongraphic studies, so ATAs between vital rates could,
in principle, be studied. To our knowledge, this has not yet been done.

%Differences with the original, analytic version of the MCT
%
%***DAN: I am not so comfortable with this because I have insufficient 
%knowledge of the original MCT. I also do not want to devote the time
%to learning it right now because of other commitments. I've reworded, 
%trying not to change the meaning. Jasmin, how confident are you in 
%these statements? We will also want to try to get Steve Ellner to check 
%them.
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic appoach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic appproach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analtic approaches provide.
It may be possible to extend the mathematics of the original MCT to consider ATAs.

>>>>

%G4
To achieve goal G4 from the Introduction, i.e., to provide an empirical example of species 
coexistence which highlights the role ATAs can play, we use a laboratory chemostat
system of freshwater diatoms explored by \cite{Descamps_2005}.
Those authors used measurements of growth rates of \emph{Fragilaria crotonensis} and 
\emph{Cyclotella pseudostelligera} over a range of temperatures to parametrize a resource
competition model. The model is 
\begin{align}
\frac{dS}{dt} &= D(S_0 - S) - Q_1x_1 \frac{V_1 S}{K_1 + S} - Q_2 x_2 \frac{V_2 S}{K_2 + S} \\
\frac{dx_j}{dt} &= x_j \frac{V_j S}{K_j+S}-D x_j, 
\end{align}
for $j=1,2$, where $S$ is a extracellular silicate concentration in the chemostat, $x_1$ is the population density
of \emph{F. crotonensis} and $x_2$ is that of \emph{C. pseudostelligera}, 
$D$ is the chemostat outflow rate, $S_0$ is concentration of silicates in the chemostat
inflow, the $Q_j$ are amounts of silicate per cell, the $V_j$ are maximum population growth rates, and
the $K_j$ are half-saturation constants for the dependence of growth rates on nutrient concentration.
The temperature-dependent parameters $Q_j$, $V_j$, and $K_j$ were obtained by 
\cite{Descamps_2005} through batch experiments and curve fitting or interpolation.
\cite{Descamps_2005} showed that permanent coexistence of the two species occurs when 
temperature fluctuates periodically, 
$\theta(t)=\theta_0 +  a\sin(2\pi t/P)$, 
with parameters $\theta_0 = 18^\circ$C, $a = 6^\circ$C, $P = 60$d. 
\cite{Ellner_2016, Ellner_2019} broke down coexistence into contributing mechanisms
according to (\ref{eq:previous_decomp}) for these same values of
$\theta_0$, $a$ and $P$. We further decompose
coexistence according to (\ref{eq:full_decomp}), and we do so for ranges of values
of $\theta_0$, $a$ and $P$. In contrast to the lottery model case, for which simplifying assumptions
made it sufficient to decompose only the GWR of species 1, for this
example we considered $\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$ for both $i=1$, $j=2$ and for
$i=2$, $j=1$. Further details of the model setup are in SI section \ref{SI-sect:diatom_model}
and specifics of how our decomposition extends to this empirical case are in 
SI section \ref{SI-sect:Empircal_app}.

@@

%ATA specialists
Contributions of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist in some contexts. If a strong competitor is present across a
collection of habitat patches and the GWR of a weaker competitor is only positive 
in a patch in the presence of 
ATAs between the temporal fluctuations of two environmental variables in that patch,
then the weaker competitor can only persist in patches for which ATAs are present.
The weaker competitor can therefore perhaps be regarded as an ``ATA specialist'' in the landscape. EC-associations
can often boil down to associations between two environmental variables, as was assumed to be the case
for the lottery model. Mechanisms of coexistence should correspond to
means by which niche differentiation is possible. Thinking about our new mechanism of 
coexistence in terms of possible ATA specialists helps highlight this connection. Likewise,
because we have described ATA exclusion as well as ATA rescue, it is possible to 
imagine a scenario in which a weaker competitor can only exist in habitat patches 
for which ATAs are absent, and is hence a specialist on scenarios with symmetric
tail associations of environmental variables (an ``STA specialist''). Additional 
research will be needed to assess to what degree these possibilities are realized 
in nature.

@@

%E and vital rates
Whereas the ``$E$'' of ``$EC$-covariance'' is often conceptualized as ``environment'', in fact it is a 
species response to the environment. Because a directly measurable aspect of the 
environment and a species response to it may be nonlinearly related, it may be 
important, in future work, 
to also study the nature of tail associations between species vital rates.
In the lottery model, it was assumed that fecundity of species $i$ tracks the
environment directly, $B_i(t)=E_i(t)$. But vital rates such as fecundity are 
more likely related to actual environmental variables according to a nonlinear
function which may often be threshold-like or sigmoidal, and thus the $E_i=B_i$
of the lottery model is actually better conceptualized as a species response
than as the environment, directly.
Likewise, one of the choices of $E_i$ considered by \cite{Ellner_2016} for the
diatom model was $E_i=V_i$, the maximum population growth rate, clearly a response
to the environment rather than a direct environmental measurement. 
It was demonstrated elsewhere that nonlinear transformations can produce or alter 
asymmetries of tail association \citep{Ghosh_2020_AER,Walter_2022}.
Thus the nature of ATAs between directly measured environmental variables may differ from the
nature of ATAs between the population responses which actually matter for coexistence.
Vital rates are often measured directly in demographic studies, so ATAs between vital rates could,
in principle, be studied. To our knowledge, this has not yet been done.

%Differences with the original, analytic version of the MCT
%
%***DAN: I am not so comfortable with this because I have insufficient 
%knowledge of the original MCT. I also do not want to devote the time
%to learning it right now because of other commitments. I've reworded, 
%trying not to change the meaning. Jasmin, how confident are you in 
%these statements? We will also want to try to get Steve Ellner to check 
%them.
Our approach to exploring the influence of ATAs on coexistence relies on the computational
extension of the MCT of \cite{Ellner_2019},
and our results help demonstrate strengths of that approach.
Storage effects as computed using the original, analytic approach to MCT 
\citep{Chesson_1994,Chesson_2000} should equal our term $\Delta^{[E \nat C]}$, 
which we refer to as the contributions to coexistence of correlation \emph{per se}.
This is because the weak-noise/Taylor approximation approach used in the original 
MCT neglects the effects of ATAs when higher-order terms
in Taylor expansions are dropped. The term $\Delta^{(E \shp C)}$, which our theory 
inherits from \cite{Ellner_2016}, is also dropped. These observations do not diminish 
the major strengths of the 
original, analytic approach, which are well known and also elaborated by \cite{Ellner_2019};
one such strength is the generality that analytic approaches provide.
It may be possible to extend the mathematics of the original MCT to consider ATAs.
